gazeHMM: Parameter recovery simulation

### 2020-05-06 simulation/preregistration_parameter_recovery.Rmd
add simulation method for gamma distribution

The parameter recovery study for the HMM is divided into two parts: In the first part, we will vary the parameters of the model and estimate the recovery. In the second part, the parameters of the model will be fixed. Instead, we will manipulate the size of the sample generated by the model level, the noise added to the generated data, the starting values for the HMM, and the amount of missing data.

## Variable parameters

First, we will repeatedly generate data with the HMM and the parameters displayed in Table 2. Second, the HMM estimates the parameters from the generated data. Third, we will compare the estimated parameters with the true parameters. We will treat the parameter recovery as satisfactory when the 95% confidence bands include the true parameter values.

@@

|State|Parameter|Min.|Max.|Fixed|
|-----|---------|----|----|-----|
|1-4|$\pi$|-|-|$1/k$|
|1-4|$a_{i=j}$|0.01|0.99|0.9|
|1-4|$a_{i \neq j}$|0.01|0.99|0.1|
|1|$\alpha_{vel}$|1|5|3|
|1|$\beta_{vel}$|0.5|1|0.75|
|1|$\alpha_{acc}$|1|5|3|
|1|$\beta_{acc}$|0.05|0.5|0.15|
|2|$\alpha_{vel}$|1|5|3|
|2|$\beta_{vel}$|5|15|10|
|2|$\alpha_{acc}$|1|5|3|
|2|$\beta_{acc}$|1|5|3|
|2|$\mu$|-|-|0|
|2|$\kappa$|0.1|10|1|
|3|$\alpha_{vel}$|1|5|3|
|3|$\beta_{vel}$|1|5|3|
|3|$\alpha_{acc}$|1|3|2|
|3|$\beta_{acc}$|1|3|2|
|3|$\mu$|-|-|$\pi$|
|3|$\kappa$|0.1|10|1|
|4|$\alpha_{vel}$|1|5|3|
|4|$\beta_{vel}$|1|2|1.5|
|4|$\alpha_{acc}$|1|5|3|
|4|$\beta_{acc}$|0.05|0.25|0.15|
|4|$\mu$|-|-|0|
|4|$\kappa$|0.1|10|1|

## Scenario 2: Sample size and noise

In this scenario, the parameters of the HMM are fixed to the values shown in Table 2. For HMMs with two, three, and four states, we will repeatedly generate data with the model for 500, 2500, and 10.000 samples. In addition, gaussian white noise with paramters $\sigma_{vel; acc} \in \{1, 5, 25\}$ and $\kappa_{angle} \in \{0.1, 1, 10\}$ will be added to the data. This will results in 27 ($5 \times 3 \times 3$) conditions with *D* repetitions each.

##

>>>>

To estimate parameter recovery of the HMM, we repeatedly generate data with the model under different conditions and a different number of states. For every condition and number of states $k \in \{2, 3, 4\}$, the model will generate $D = 100$ data sets with parameters evenly distributed on a given interval. The same model is then applied to estimate the parameters from the generated data. Afterwards, the estimated parameters are compared with the "true" parameters that generated the data. We treat a parameter as successfully recovered when the 95% confidence bands of the estimated parameters include the true parameters. We further require a 90% median rate of correct classifications across the data sets for the model to be satisfactorily accurate.

The HMM will always start with a uniform distribution to estimate the initial state and state transition probabilites. To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with the true parameter as the shape parameter and a scale parameter of $\beta = 1$. Mean parameters of the von-Mises distribution will always start at their true value.

In the first part of the simulation study, we will estimate parameter recovery under ideal condititions. These imply an average sample size of $N = 2500$ samples per data set, no noise added to the data, and no missing samples. The parameters that we will use to generate the data are displayed in Table 2. We will vary one parameter at a time on the given interval and set the other parameters to their fixed value. This simulation will result in $D_{total} = 100 \times 3 \times 20 = 6000$ recoveries.

In the second part, we will vary the sample size of the generated data and the amount of noise added to it. For sample sizes of $N \in \{500, 2500, 10000\}$, we will generate data sets and add white noise. To velocity and acceleration, we will add gaussian noise with $\mu = 0$ and $\sigma \in [1,25]$. Noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ will be added to angle. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

In the third part, we will increase the variation in the starting values used for parameter estimation. For the shape, scale, and concentration parameters, we will increase the scale parameters of the gamma distributions to $\beta \in \{2, 4, 8\}$ all at the same time. The sample size will be $N = 2500$ and no noise will be added to the data. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

In the last part, we will set intervals of the generated data to be missing. The size of the missing data interval will be $m \in [1,100]$ and the number of intervals will be $n_{miss} \in \{1, 3, 5\}$. The sample size will be $N = 2500$ and no noise will be added to the data. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

@@

|State|Parameter|Interval|Fixed|
|-----|---------|--------|-----|
|1-4|$p$*|-|$1/k$|
|1-4|$a_{i=j}$|[.01,.99]|0.9|
|1-4|$a_{i \neq j}$|$(1-a_{i=j})/(k-1)$|$0.1/(k-1)$|
|1|$\alpha_{vel}$|[1,5]|3|
|1|$\beta_{vel}$|[0.5,1]|0.75|
|1|$\alpha_{acc}$|[1,5]|3|
|1|$\beta_{acc}$|[0.05,0.5]|0.15|
|1|$a$*|-|-|
|1|$b$*|-|-|
|2|$\alpha_{vel}$|[1,5]|3|
|2|$\beta_{vel}$|[5,15]|10|
|2|$\alpha_{acc}$|[1,5]|3|
|2|$\beta_{acc}$|[1,5]|3|
|2|$\mu$*|-|0|
|2|$\kappa$|$1/[0.1,10]$|1|
|3|$\alpha_{vel}$|[1,5]|3|
|3|$\beta_{vel}$|[1,5]|3|
|3|$\alpha_{acc}$|[1,3]|2|
|3|$\beta_{acc}$|[1,3]|2|
|3|$\mu$*|-|$\pi$|
|3|$\kappa$|$1/[0.1,10]$|1|
|4|$\alpha_{vel}$|[1,5]|3|
|4|$\beta_{vel}$|[1,2]|1.5|
|4|$\alpha_{acc}$|[1,5]|3|
|4|$\beta_{acc}$|[0.05,0.25]|0.15|
|4|$\mu$*|-|0|
|4|$\kappa$|$1/[0.1,10]$|1|

*Note*. Parameters marked with * will not be varied but always set to their fixed values. Parameters *a* and *b* (the minimum and maximum of the uniform distribution) are always obtained from the data.
